\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-16 NAR_DRAFT.tex
abstract. probably worse

\begin{abstract}
The accumulation of somatic mutations is an inevitable consequence of tumor development.
Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in treatment resistance, and the value of tumor heterogeneity as a diagnostic biomarker.
The identification of low frequency mutations from Next Generation Sequencing data remains challenging.
In general the ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
\rngcomment{A little too much intro in abstract.}
This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
\batcave uses the mutation profile and allele frequency spectrum of high-confidence mutations to estimate the data generating distribution giving rise to variants.
We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
\rngcomment{Need a sentence about what new analysis this enables. How does \batcave (at least help) solve the problem you raised earlier?}
We provide an R package implementing the \batcave algorithm for use with MuTect output.
\end{abstract}

>>>>

\begin{abstract}
%The accumulation of somatic mutations is an inevitable consequence of tumor development.
Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in treatment resistance, and the value of tumor heterogeneity as a diagnostic biomarker.
The ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
% \rngcomment{A little too much intro in abstract.}
% This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
\batcave uses the mutation profile and allele frequency spectrum of high-confidence mutations to estimate the data generating distribution giving rise to variants.
We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
The algorithm performs consistently as sequencing depth increases, enabling deeper sequencing and the identification lower frequency variants without generating excessive false positive calls.
\rngcomment{Need a sentence about what new analysis this enables. How does \batcave (at least help) solve the problem you raised earlier?}
We provide an R package implementing the \batcave algorithm for use with MuTect output.
\end{abstract}
